Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-08 writeup_and_notes/male_dependent_drive.tex
fixed tupos, modified the bistabilty section

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is competed only after fertilization \citep{Masui_book}, 
creating ample opportunity for interaction between the sperm and female meiotic machinery. 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by selfish sperm systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals, 
	female meiosis is complete upon ovulation (that is fertilization is not required for female meiosis \citep[see table 1 of ref 
][]{Masui_book}).

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg  in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype. To depict a case
of a self-promoting meiotic driver,  we modify this standard model such that the driver is only effective when fertilized by a sperm carrying that allele (see Figure
\ref{Eggsperm_2_allele_cartoon}). 
\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$? \citep[see ][ for
        discussion]{Buckler1999}. }. %PUT THIS BACK IN
We then identify the conditions allowing the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors.

For comparative purposes, we present the standard drive model in equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious in both sexes but fully recessive 
	(i.e. homozygous fitness equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$) a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. The
        parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A., and where it can fix in the green
        and orange areas \ref{Invasion_space}B. When the allele is
        maintained as a polymorphism it providing an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
	a self-promoting driver specifically creates homozygote
        zygotes when it drives in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

The positive frequency dependence of a self-promoting driver can
induce bistability in its spread, i.e. there is a critical frequency
above which it can spread quickly to fixation and below which it moves towards loss (an
unstable equilibrium frequency). In 
the appendix we devlop a simple
approximation to this bistability 
This bistability occurs in the region of parameter space in Figure
\ref{Invasion_space}. In Supplementary Figure
\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
show that the unstable equilibrium frequency is high across much of
this parameter space, and so in practice the self promoting allele
only invade in the orange space. (In the appendix we develop a simple
approximation to the unstable equilibrium frequency
\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
\ref{Bistab_homozyg_cost_fig}).
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
strongly constrains the evolution of a self-promoting driver. 
In fact with any heterozygous fitness cost, a rare self-promoting
driver cannot spread deterministically. However, this case also displays bistability -- 
if $s < d - 1/2+s_h(3-2d)/2$ approximately, it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
observed as stable polymorphisms in natural populations.

>>>>

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
creating ample opportunity for interaction between the sperm and female meiotic machinery. 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is certainly not the case that animals are mechanistically incapable of circumventing this requirement.
The stage at which female meiosis requires fertilization varies widely across taxa, and in a few animals, 
	female meiosis is complete upon ovulation (that is, fertilization is not required for female meiosis \citep[see table 1 of ref 
][]{Masui_book}).

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg  in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype. To depict a case
of a self-promoting meiotic driver,  we modify this standard model such that the driver is only effective when fertilized by a sperm carrying that allele (see Figure
\ref{Eggsperm_2_allele_cartoon})
\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$? \citep[see ][ for
        discussion]{Buckler1999}. }. %PUT THIS BACK IN
We then identify the conditions allowing the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors.

For comparative purposes, we present the standard drive model in equations \ref{eqn.A1}-\ref{eqn.A3}, 
	and refer readers to the previous literature on this model \citep[e.g. ][]{Prout1973,Ubeda2004} for additional results. 
In this case, assuming the driving allele is deleterious in both sexes but fully recessive 
	(i.e. homozygous fitness equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$) a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. The
        parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A., and where it can fix in the green
        and orange areas \ref{Invasion_space}B. When the allele is
        maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
	a self-promoting driver specifically creates homozygote
        zygotes when it drives. %in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

%I SUBSTANTIALLY REWROTE THIS
\yb{
The positive frequency dependence of a self-promoting driver can
induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in EQUATION XXX),
there is  an unstable equilibrium frequency (approximated in EQUATION XXX)  
	above which it can spread quickly to fixation and below which it moves towards loss 
	(see Figure \ref{Bistab_homozyg_cost_fig} for exact results).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
This unstable equilibrium frequency is high across much of
	parameter space, likely too high to be reached by drift, 
	and therefore the more restrictive invasion criteria rather than the fixation criteria, will  likely determine the fate of self promoting drivers. 
}

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-h_s$)
further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
driver cannot spread deterministically. 
However, this case also displays bistability -- 
if $s < d - 1/2+s_h(3-2d)/2$ approximately, it will fix deterministically if its 
	frequency is greater than a threshold, $f_B^*$ [Equation \eqref{thresh2}, and \yb{FIGURES SXX}].
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that, if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
observed as stable polymorphisms in natural populations.
